---
title: "Ongo Science data cleaning and feature engineering for user churn modeling"
author: "Rachel Wehr, PhD; Health Data Science Fellow, Insight"
output:
  pdf_document: default
  html_notebook: default
---
```{r}
library(plyr)
library(dplyr)
library(gridExtra)
library(eeptools)
library(reshape2)
library(data.table)
library(tidyverse)
library(lubridate)
require("RPostgreSQL")
require("sqldf")

```


The goal of this is to directly interface with the PostgreSQL database hosted on AWS. This provides up-to-date information about user in-app behavior. 

```{r}
# Connect to the PostgreSQL database
# save the password that we can "hide" it as best as we can by collapsing it
pw <- {
  "" # add password 
}
 
# loads the PostgreSQL driver
drv <- dbDriver("PostgreSQL")
# creates a connection to the postgres database
# note that "con" will be used later in each connection to the database
con <- {
  dbConnect(drv, dbname = "", # include dbname
                 host = "", port = 0000, #include host and port number
                 user = "", password = pw) # include user, and pw from above
}

#rm(pw) # removes the password
 
# check for the existance of the tables to pull in
dbExistsTable(con, "User")
dbExistsTable(con, "UserLog")
dbExistsTable(con, "DataWorkoutRun")
dbExistsTable(con, "TrackSessionComplete")
dbExistsTable(con, "TrackEnrollment")
dbExistsTable(con, "ProfileTimelineItem")
dbExistsTable(con, "SubscriptionPlan")
dbExistsTable(con, "SubscriptionReceipt")
dbExistsTable(con, "DataActivity")

# be sure to disconnect from connection
#dbDisconnect(con)
```

Next step is to set the working directory of choice and to read the tables from the databases in from PostgreSQL
```{r}
setwd('') #set the output

# Pull in the tables directly from the PostgreSQL database
user <- dbReadTable(con, c("User"))
userLog <- dbReadTable(con, c("UserLog"))
dwRun <- dbReadTable(con, c("DataWorkoutRun"))
trackEnrollment <- dbReadTable(con, c("TrackEnrollment"))
trackSessionComplete <- dbReadTable(con, c("TrackSessionComplete"))
trackEnrollment <- dbReadTable(con, c("TrackEnrollment"))
trackComplete <- dbReadTable(con, c("TrackComplete"))
profTimelineitem <- dbReadTable(con, c("ProfileTimelineItem"))
subscriptionPlan <- dbReadTable(con, c("SubscriptionPlan")) # gives indication if Monthly or 12 months
subscription <- dbReadTable(con, c("SubscriptionReceipt")) # can then link to the ID
dataActivity <- dbReadTable(con, c("DataActivity")) 

#dbDisconnect(con)

```


```{r}
# Pull in USER INFORMATION, including demographics
user2 <- data.frame("userID" = user$X_id, "birthday" = as.character(user$dateOfBirth), "gender" = user$gender)
# convert to get the bday in date and extract ages
temp <- ymd_hms(user2$birthday)
user2$bday <- as.Date(temp)
ages <- age_calc(na.omit(user2$bday), units = "years")
user2$age[!is.na(user2$bday)] <- ages


#  ==============
### The following can be updated to lengthen or shorten the times of interest to investigate
firstdate <- "2019-08-04" # day of app launch
lastdate <- "2019-09-24" # this was the last date that I analyzed data for
minnumdays <- 14 # change to 14 for most analyses, 1 for EDA histogram graphs
#  ==============

userLog2 <- data.frame("userID" = userLog$user, "logtype" = userLog$type, "date" = as.Date(userLog$createdAt))

# limit only to those between the two date of interest (App launch, and final analysis date)
userLog2 <- userLog2[which(userLog2$date >= firstdate & userLog2$date <= lastdate),]

# Determine the first login date for each user
firstLogin <- userLog2 %>%
    group_by(userID) %>%
    arrange(date) %>%
    slice(1L)

userInfoAll <- full_join(user2,firstLogin,by="userID")
# contains the information ONLY for those that had a login during August
userInfoUnion <- right_join(user2,firstLogin,by="userID")


# ========================================================================================================
### Engineer data frame with a date for each column, userID row, and # observations for logins
# ========================================================================================================

#### Create temporary data frame with count of logins by date for each unique user
  temptable <- count(userLog2, userID, date) # gives count for each available date per user
  temptable$date <- as.Date(as.character(temptable$date))
  loginInfo <- dcast(temptable, userID ~ date)
  loginInfo[is.na(loginInfo)] <- 0
# combine this with the demographic info
  loginInfo2 <- right_join(userInfoUnion,loginInfo,by="userID")

  
# Sum together the login info for the columns with dates for each user for total logins in last month
  totaldays <- length(which(grepl("20", colnames(loginInfo2))))
  for (i in 1:length(loginInfo2$userID)){
    loginInfo2$totalLogins[i] <- sum(loginInfo2[i,8:(7+totaldays)])}
  
 # rename first active date for each user
 names(loginInfo2)[names(loginInfo2) == 'date'] <- 'firstlogindate'
 # Specify the last date in the dataset to set parameter for time with app and provide columns with # days user has had app
 lastDate <- max(userLog2$date)
 loginInfo2$dayswithApp <- age_calc(na.omit(loginInfo2$firstlogindate), as.Date(lastDate), units="days")
 loginInfo2$dayswithAppNumeric <- as.numeric(loginInfo2$dayswithApp) + 1

```

FEATURE #1: Number of logins during week 1
The next section Builds on loginInfo7 to determine the user-specific behavior during EACH USER's first seven days with the app.
```{r}

#### First, we identify which subjects we will include; for the model, we are removing subjects that have 14 days or less of data
## This is because 1) users of fitness apps do not use the app every day, but might have specific weekly patterns in exercise, and
## 2) I had only about 52 days of data, so I chose 7 days to train the model, and decided that we need at least one additional week 
##    for each user to actively predict future behavior.
 loginInfo7 <- loginInfo2[which(loginInfo2$dayswithAppNumeric > minnumdays),]

# Figuring out the number of logins within the first 7 days & number of days logged in in trial period
# create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumns <- which(grepl("20", colnames(loginInfo7)))
 allDates <- colnames(loginInfo7[IndexColumns])
 
 # adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(loginInfo7$firstlogindate)){
      loginInfo7$startvec[i] <- which(allDates == loginInfo7$firstlogindate[i])# true date 1 for user in each row
      loginInfo7$lastdatevec[i] <- length(allDates) # true last date logged 
        }
  # gives the column for the last date in the trial period 
loginInfo7$endvec <- loginInfo7$startvec + 6 # for 7th day after 1st login
loginInfo7$startcol <- IndexColumns[loginInfo7$startvec] # gives column for date of 1st login
loginInfo7$endcol <- IndexColumns[loginInfo7$endvec] # gives column for date of last date in week 1
loginInfo7$lastdatecol <- IndexColumns[loginInfo7$lastdatevec] # gives column date of last day


# gives the number of total login ins for the user's first week
# also gives the number of active days for the user's first week
# contains info for logins following week 1 as well to help indicate churn
for (i in 1:length(loginInfo7$firstlogindate)){
  mycols <- loginInfo7$startcol[i]:loginInfo7$endcol[i]
  postwk1cols <- (loginInfo7$endcol[i]+1):(loginInfo7$lastdatecol[i])
  loginInfo7$NumfirstWeeklogins[i] <- sum(loginInfo7[i, mycols])
  loginInfo7$Wk1ActiveDays[i] <- rowSums(loginInfo7[i,mycols] != 0)
  loginInfo7$NumPostwk1Logins[i] = sum(loginInfo7[i, postwk1cols])
  loginInfo7$Postwk1ActiveDays[i] <- rowSums(loginInfo7[i,postwk1cols] != 0)
} 
```


I am using the in-app login data engineered above to define our levels of churn. For this analysis, we are defining churn in two ways to build two separate models that will predict churn by two different threshholds. This next section binarizes the first definition of churn, which is an individual who has logged in MORE THAN ONE time following the first seven days. 
```{r}
# determine post week 1 churn (churners = 1, those logged in at least twice = 0)
loginchurners <- which(loginInfo7$NumPostwk1Logins < 2) # index for churners
# Incorporate into loginInfo7 data frame
loginInfo7$loginchurnPostwk1 <- 0
loginInfo7$loginchurnPostwk1[loginchurners] <- 1


# Normalize the number of logins to number of dates with app
loginInfo7$NumPostDays <- loginInfo7$dayswithAppNumeric - 7
loginInfo7$loginsNormPostDays <- loginInfo7$NumPostwk1Logins/loginInfo7$NumPostDays #

### NEXT: add some information such as what the users DID during their logins 
  # define power users (likely to be coaches?) those that logged in > 250 times
  powerUsers <- which(loginInfo7$totalLogins > 250)
  loginInfo7 <- loginInfo7[-powerUsers,]
```

FEATURE #2: Number of GPS-enabled runs during week 1
```{r}
# Pull the information from the dwRun dataframe above
dwRun2 <- data.frame("userID" = dwRun$owner, "inORout" = dwRun$type, "track" = dwRun$track, "sessionTitle" = dwRun$sessionTitle, "trackTitle" = dwRun$trackTitle, "date" = as.Date(dwRun$startDate))

# limit only to between specific dates, and those instances that were SOLO runs, NOT included in session/track
dwRun2 <- dwRun2[which(dwRun2$date >= firstdate & dwRun2$date <= lastdate),] 
# adding this next step removes any runs that were part of the session/tracks
dwRun2 <- dwRun2[which(is.na(dwRun2$sessionTitle) == FALSE),]
dwRun2$occur <- 1

### Quick view of the popularity of tracks and sessions within the last month
aggregate(occur ~ trackTitle + date, data = dwRun2, sum)
trackTitleoccurances <- aggregate(occur ~ trackTitle, data = dwRun2, sum)
trackTitleoccurances[order(trackTitleoccurances$occur, decreasing=TRUE),] # shows popularity of tracks in last month
trackSessionoccurances <- aggregate(occur ~ sessionTitle, data = dwRun2, sum)
trackSessionoccurances[order(trackSessionoccurances$occur, decreasing=TRUE),] # shows popularity of tracks in last month

## Looking at if a user tracked a run
#### Create temporary data frame with count of runs by date for each unique user
# this is taking the dates from ALL available; not just run dates
  temptableRUN <- count(dwRun2, userID, date) # gives count for each available date per user
  temptableRUN$date <- as.Date(as.character(temptableRUN$date))
  # Cast this into a data frame to count the number of solo runs completed by date for each user
  dwRun2 <- dcast(temptableRUN, userID ~ date)

# find any dates that are available logins but did NOT have any user runs tracked
  diffColnames <- setdiff(colnames(loginInfo7), colnames(dwRun2))
  indexRUNaddDate <- diffColnames[which(grepl("20", diffColnames))]
  dwRun2$placeholder <- 0
  # rename any additional date --> will need to update if more than one date
  names(dwRun2)[names(dwRun2) == "placeholder"] <- indexRUNaddDate
  dwRun2 <- dwRun2[,order(colnames(dwRun2), decreasing=FALSE)] 
  dwRun2[is.na(dwRun2)] <- 0
  
 # Add first login date from loginInfo7 so that we can calculate the num runs logged in wk1
 tempdf <- data.frame("userID"=loginInfo7$userID, "firstlogindate"=as.character(as.Date(loginInfo7$firstlogindate)))
 dwRun3 <-  left_join(dwRun2, tempdf, by = "userID")
 
 # create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumnsRUN <- which(grepl("20", colnames(dwRun3)))
 allDatesRUN <- colnames(dwRun3[IndexColumnsRUN])
 # Ensure that you have the same amount of dates in this as you do the loginInfo7 
 length(allDates) == length(allDatesRUN)
 
 ### remove anyone that does not have a valid start date; this is a user not listed in loginInfo7
 ## likely because they were removed for not having enough data dat %>% drop_na("B")
 #allDatesRUN <- allDatesRUN %>% drop_na("firstlogindate")
 dwRun3 <- dwRun3 %>% drop_na("firstlogindate")  
 
 # adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(dwRun3$userID)){
      dwRun3$startvec[i] <- which(allDatesRUN == dwRun3$firstlogindate[i])# for user in each row
      dwRun3$lastdatevec[i] <- length(allDatesRUN)
 }

  # gives the column for the last date in the trial period 
dwRun3$endvec <- dwRun3$startvec + 6
dwRun3$startcol <- IndexColumnsRUN[dwRun3$startvec]
dwRun3$endcol <- IndexColumnsRUN[dwRun3$endvec]
dwRun3$lastdatecol <- IndexColumnsRUN[dwRun3$lastdatevec]

# gives the number of total runs logged ins for the user's first week
# also gives the number of active run days for the user's first week
# contains info for runs following week 1 as well to help indicate churn
for (i in 1:length(dwRun3$firstlogindate)){
  myrows <- dwRun3$startcol[i]:dwRun3$endcol[i]
  postwk1rows <- (dwRun3$endcol[i]+1):(dwRun3$lastdatecol[i])
  dwRun3$NumfirstWeekRuns[i] <- sum(dwRun3[i, myrows])
  dwRun3$NumPostwk1Runs[i] = sum(dwRun3[i, postwk1rows])
} 
  
### ADD NumfirstWeekRuns and Numpostwk1Runs to loginInfo7
tempdf <- data.frame("userID"=dwRun3$userID, "NumPostwk1Runs"=dwRun3$NumPostwk1Runs, "NumfirstWeekRuns" = dwRun3$NumfirstWeekRuns)
loginInfo7 <-  left_join(loginInfo7, tempdf, by = "userID")

```

FEATURE #3: Number of tracks that the user enrolled in during week 1
```{r}
## Looking at track enrollment information
trackEnrollment2 <- data.frame("userID" = trackEnrollment$user, "track" = trackEnrollment$track, "date" = as.Date(trackEnrollment$createdAt))
trackEnrollment2$enrollOccur <- 1

## Limit to instances between the dates of interest
trackEnrollment2 <- trackEnrollment2[which(trackEnrollment2$date >= firstdate & trackEnrollment2$date <= lastdate),]

#### Create data frame with count of logins by date for each unique user for enrollment
# this is taking the dates from ALL available; not just enrollment dates
  temptableTE <- count(trackEnrollment2, userID, date) # gives count for each available date per user
  temptableTE$date <- as.Date(as.character(temptableTE$date))
  trackEnrollment2 <- dcast(temptableTE, userID ~ date)
  # find any dates that are available logins but did NOT have any user enrollments; if no dates, comment out
    #diffColnames <- setdiff(colnames(loginInfo7), colnames(trackEnrollment2))
    #indexTEaddDate <- diffColnames[which(grepl("20", diffColnames))]
    #trackEnrollment2$placeholder <- 0
  # rename any additional date --> will need to update if more than one date
    #names(trackEnrollment2)[names(trackEnrollment2) == "placeholder"] <- indexTEaddDate
  trackEnrollment2 <- trackEnrollment2[,order(colnames(trackEnrollment2), decreasing=FALSE)] 
  trackEnrollment2[is.na(trackEnrollment2)] <- 0
  
 # Add first login date from loginInfo7 so that we can calculate the num sessions enrolled in wk1
 tempdf <- data.frame("userID"=loginInfo7$userID, "firstlogindate"=as.character(as.Date(loginInfo7$firstlogindate)))
 trackEnrollment3 <-  left_join(trackEnrollment2, tempdf, by = "userID")
 
 # create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumnsTE <- which(grepl("20", colnames(trackEnrollment3)))
 allDatesTE <- colnames(trackEnrollment3[IndexColumnsTE])
 # Ensure that you have the same amount of dates in this as you do the loginInfo7 
 length(allDates) == length(allDatesTE)
 
 ### remove anyone that does not have a valid start date; this is a user not listed in loginInfo7
 ## likely because they were removed for not having enough data dat %>% drop_na("B")
 trackEnrollment3 <- trackEnrollment3 %>% drop_na("firstlogindate")

 # adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(trackEnrollment3$userID)){
      trackEnrollment3$startvec[i] <- which(allDatesTE == trackEnrollment3$firstlogindate[i])# for user in each row
      trackEnrollment3$lastdatevec[i] <- length(allDatesTE)
 }

  # gives the column for the last date in the trial period 
trackEnrollment3$endvec <- trackEnrollment3$startvec + 6
trackEnrollment3$startcol <- IndexColumnsTE[trackEnrollment3$startvec]
trackEnrollment3$endcol <- IndexColumnsTE[trackEnrollment3$endvec]
trackEnrollment3$lastdatecol <- IndexColumnsTE[trackEnrollment3$lastdatevec]

# gives the number of total login ins for the user's first week
# also gives the number of active days for the user's first week
# contains info for logins following week 1 as well to help indicate churn
for (i in 1:length(trackEnrollment3$firstlogindate)){
  myrows <- trackEnrollment3$startcol[i]:trackEnrollment3$endcol[i]
  postwk1rows <- (trackEnrollment3$endcol[i]+1):(trackEnrollment3$lastdatecol[i])
  trackEnrollment3$NumfirstWeekEnroll[i] <- sum(trackEnrollment3[i, myrows])
  trackEnrollment3$NumPostwk1Enroll[i] = sum(trackEnrollment3[i, postwk1rows])
} 

### ADD NumfirstWeekEnroll and Numpostwk1Enroll to loginInfo7
tempdf <- data.frame("userID"=trackEnrollment3$userID, "NumPostwk1Enroll"=trackEnrollment3$NumPostwk1Enroll, "NumfirstWeekEnroll" = trackEnrollment3$NumfirstWeekEnroll)
loginInfo7 <-  left_join(loginInfo7, tempdf, by = "userID")

```

FEATURE #4: Engineers the number of Sessions that were COMPLETED by a user during week 1
```{r}
#### Gives information for each session completed by users
trackSessionComplete2 <- data.frame("userID" = trackSessionComplete$user, "track" = trackSessionComplete$track, "trackTitle" = trackSessionComplete$trackTitle, "session" = trackSessionComplete$trackSession, 'sessionTitle' = trackSessionComplete$sessionTitle, "date" = as.Date(trackSessionComplete$createdAt))

# limit between dates of interest
trackSessionComplete2 <- trackSessionComplete2[which(trackSessionComplete2$date >= firstdate & trackSessionComplete2$date <= lastdate),]
trackSessionComplete2$completeOccur <- 1

#### Create data frame with count of logins by date for each unique user for enrollment
# this is taking the dates from ALL available; not just enrollment dates
  temptableTS <- count(trackSessionComplete2, userID, date) # gives count for each available date per user
  temptableTS$date <- as.Date(as.character(temptableTS$date))
  trackSessionComplete2 <- dcast(temptableTS, userID ~ date)
  # find any dates that are available logins but did NOT have any completed sessions
  diffColnames2 <- setdiff(colnames(loginInfo7), colnames(trackEnrollment2))
    #indexTSaddDate <- diffColnames2[which(grepl("20", diffColnames2))]
    #trackSessionComplete2$placeholder <- 0
    #names(trackSessionComplete2)[names(trackSessionComplete2) == "placeholder"] <- indexTSaddDate
  trackSessionComplete2 <- trackSessionComplete2[,order(colnames(trackSessionComplete2), decreasing=FALSE)] 
  trackSessionComplete2[is.na(trackSessionComplete2)] <- 0
  
 # Add first login date from loginInfo7 so that we can calculate the num sessions completed in wk1
 tempdf <- data.frame("userID"=loginInfo7$userID, "firstlogindate"=as.character(as.Date(loginInfo7$firstlogindate)))
 trackSessionComplete2 <-  left_join(trackSessionComplete2, tempdf, by = "userID")
 
 # create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumnsTS <- which(grepl("20", colnames(trackSessionComplete2)))
 allDatesTS <- colnames(trackSessionComplete2[IndexColumnsTS])
 # Ensure that you have the same amount of dates in this as you do the loginInfo7 
 length(allDates) == length(allDatesTS)
 
 ### remove anyone that does not have a valid start date; this is a user not listed in loginInfo7
 ## likely because they were removed for not having enough data dat %>% drop_na("B")
 trackSessionComplete3 <- trackSessionComplete2 %>% drop_na("firstlogindate")
 
 # adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(trackSessionComplete3$userID)){
      trackSessionComplete3$startvec[i] <- which(allDatesTS == trackSessionComplete3$firstlogindate[i])# for user in each row
      trackSessionComplete3$lastdatevec[i] <- length(allDatesTS)
 }

  # gives the column for the last date in the trial period 
trackSessionComplete3$endvec <- trackSessionComplete3$startvec + 6
trackSessionComplete3$startcol <- IndexColumnsTS[trackSessionComplete3$startvec]
trackSessionComplete3$endcol <- IndexColumnsTS[trackSessionComplete3$endvec]
trackSessionComplete3$lastdatecol <- IndexColumnsTS[trackSessionComplete3$lastdatevec]

# gives the number of total login ins for the user's first week
# also gives the number of active days for the user's first week
# contains info for logins following week 1 as well to help indicate churn
for (i in 1:length(trackSessionComplete3$firstlogindate)){
  myrows <- trackSessionComplete3$startcol[i]:trackSessionComplete3$endcol[i]
  postwk1rows <- (trackSessionComplete3$endcol[i]+1):(trackSessionComplete3$lastdatecol[i])
  trackSessionComplete3$NumfirstWeekSess[i] <- sum(trackSessionComplete3[i, myrows])
  trackSessionComplete3$NumPostwk1Sess[i] = sum(trackSessionComplete3[i, postwk1rows])
} 

### ADD NumfirstWeekEnroll and Numpostwk1runs to loginInfo7
tempdf <- data.frame("userID"=trackSessionComplete3$userID, "NumPostwk1Sess"=trackSessionComplete3$NumPostwk1Sess, "NumfirstWeekSess" = trackSessionComplete3$NumfirstWeekSess)
loginInfo7 <-  left_join(loginInfo7, tempdf, by = "userID")

```

FEATURE #5: Engineering the number of social/community interactions during week 1. 
This includes likes, comments, news story posts, upvotes
```{r}
## Include any social media posts [question, question upvote, answer, answer upvote, newsfeedstory, newsfeedstory comment, newsfeedstorylike]
profTimelineitem2 <- data.frame("userID" = profTimelineitem$user, "objType" = profTimelineitem$objType, "date" = as.Date(profTimelineitem$createdAt))
# limit between time periods of interest
profTimelineitem2 <- profTimelineitem2[which(profTimelineitem2$date >= firstdate & profTimelineitem2$date <= lastdate),]
profTimelineitem2$socialmedia <- 1

## Separate out comments/questions/answers from upvotes/likes
# remove tracksessioncomplete, trackcomplete, dataworkoutrun from profTimelineitem
remove.list <- paste(c("TrackSessionComplete", "TrackComplete", "DataWorkoutRun"), collapse = '|')
profTimelineitem3 <- profTimelineitem2 %>% filter(!grepl(remove.list, objType)) 

#### Create data frame with count of logins by date for each unique user for social media engagement
# this is taking the dates from ALL available; not just SM dates
  temptableSM <- count(profTimelineitem2, userID, date) # gives count for each available date per user
  temptableSM$date <- as.Date(as.character(temptableSM$date))
  profTimelineitem3 <- dcast(temptableSM, userID ~ date)
  # find any dates that are available logins but did NOT have any social media engagement
  diffColnames2 <- setdiff(colnames(loginInfo7), colnames(profTimelineitem3))
    #indexSMaddDate <- diffColnames2[which(grepl("20", diffColnames2))]
    #profTimelineitem3$placeholder <- 0
  # rename any additional date --> will need to update if more than one date
  # Is not required for social media engagement dates
    #names(profTimelineitem3)[names(profTimelineitem3) == "placeholder"] <- indexSMaddDate
  profTimelineitem3 <- profTimelineitem3[,order(colnames(profTimelineitem3), decreasing=FALSE)] 
  profTimelineitem3[is.na(profTimelineitem3)] <- 0
  
 # Add first login date from loginInfo7 so that we can calculate the num sessions completed in wk1
 tempdf <- data.frame("userID"=loginInfo7$userID, "firstlogindate"=as.character(as.Date(loginInfo7$firstlogindate)))
 profTimelineitem3 <-  left_join(profTimelineitem3, tempdf, by = "userID")
 
 # create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumnsSM <- which(grepl("20", colnames(profTimelineitem3)))
 allDatesSM <- colnames(profTimelineitem3[IndexColumnsSM])
 # Ensure that you have the same amount of dates in this as you do the loginInfo7 
 length(allDates) == length(allDatesSM)
 
 ### remove anyone that does not have a valid start date; this is a user not listed in loginInfo7
 ## likely because they were removed for not having enough data dat %>% drop_na("B")
 profTimelineitem3 <- profTimelineitem3 %>% drop_na("firstlogindate")
 
# adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(profTimelineitem3$userID)){
      profTimelineitem3$startvec[i] <- which(allDatesSM == profTimelineitem3$firstlogindate[i])# for user in each row
      profTimelineitem3$lastdatevec[i] <- length(allDatesSM)
 }

  # gives the column for the last date in the trial period 
profTimelineitem3$endvec <- profTimelineitem3$startvec + 6
profTimelineitem3$startcol <- IndexColumnsSM[profTimelineitem3$startvec]
profTimelineitem3$endcol <- IndexColumnsSM[profTimelineitem3$endvec]
profTimelineitem3$lastdatecol <- IndexColumnsSM[profTimelineitem3$lastdatevec]

# gives the number of total login ins for the user's first week
# also gives the number of active days for the user's first week
# contains info for logins following week 1 as well to help indicate churn
for (i in 1:length(profTimelineitem3$firstlogindate)){
  myrows <- profTimelineitem3$startcol[i]:profTimelineitem3$endcol[i]
  postwk1rows <- (profTimelineitem3$endcol[i]+1):(profTimelineitem3$lastdatecol[i])
  profTimelineitem3$NumfirstWeekSM[i] <- sum(profTimelineitem3[i, myrows])
  profTimelineitem3$NumPostwk1SM[i] = sum(profTimelineitem3[i, postwk1rows])
} 

### ADD NumfirstWeekSM and Numpostwk1SM to loginInfo7
tempdf <- data.frame("userID"=profTimelineitem3$userID, "NumPostwk1SM"=profTimelineitem3$NumPostwk1SM, "NumfirstWeekSM" = profTimelineitem3$NumfirstWeekSM)
loginInfo7 <-  left_join(loginInfo7, tempdf, by = "userID")

```

FEATURE #6: Active logins during week 1.
This is also used to binarize active churn for model # 2. 
```{r}
## Look at ANY ACTIVITY at all pre/post week 1 for active logging so that I can 
## Identify chuners as someone who did any meaningful activity post wk 1
activeLogins <- data.frame("userID" = profTimelineitem$user, "objType" = profTimelineitem$objType, "date" = as.Date(profTimelineitem$createdAt))

# limit between time periods of interest
activeLogins <- activeLogins[which(activeLogins$date >= firstdate & activeLogins$date <= lastdate),]
activeLogins$activeLogin <- 1

#### Create data frame with count of logins by date for each unique user for all active logins
# this is taking the dates from ALL available
  temptableAL <- count(activeLogins, userID, date) # gives count for each available date per user
  temptableAL$date <- as.Date(as.character(temptableAL$date))
  activeLogins2 <- dcast(temptableAL, userID ~ date)
  # find any dates that are available logins but did NOT have any active logins
  diffColnames2 <- setdiff(colnames(loginInfo7), colnames(activeLogins2))
  
  # rename any additional date --> will need to update if more than one date
  activeLogins2 <- activeLogins2[,order(colnames(activeLogins2), decreasing=FALSE)] 
  activeLogins2[is.na(activeLogins2)] <- 0
  
 # Add first login date from loginInfo7 so that we can calculate the active logins in wk 1
 tempdf <- data.frame("userID"=loginInfo7$userID, "firstlogindate"=as.character(as.Date(loginInfo7$firstlogindate)))
 activeLogins2 <-  left_join(activeLogins2, tempdf, by = "userID")
 
 # create a vector that gives the column index for any column name that begins with "20" for "2019-01-01"
 IndexColumnsAL <- which(grepl("20", colnames(activeLogins2)))
 allDatesAL <- colnames(activeLogins2[IndexColumnsAL])
 # Ensure that you have the same amount of dates in this as you do the loginInfo7 
 length(allDates) == length(allDatesAL)
 
 ### remove anyone that does not have a valid start date; this is a user not listed in loginInfo7
 ## likely because they were removed for not having enough data dat %>% drop_na("B")
 activeLogins2 <- activeLogins2 %>% drop_na("firstlogindate")

# adds 'startvec' to the data frame to tell what column to index into for first login
 for (i in 1:length(activeLogins2$userID)){
      activeLogins2$startvec[i] <- which(allDatesAL == activeLogins2$firstlogindate[i])# for user in each row
      activeLogins2$lastdatevec[i] <- length(allDatesAL)
 }

  # gives the column for the last date in the trial period 
activeLogins2$endvec <- activeLogins2$startvec + 6
activeLogins2$startcol <- IndexColumnsAL[activeLogins2$startvec]
activeLogins2$endcol <- IndexColumnsAL[activeLogins2$endvec]
activeLogins2$lastdatecol <- IndexColumnsAL[activeLogins2$lastdatevec]

# gives the number of total login ins for the user's first week
# also gives the number of active days for the user's first week
# contains info for logins following week 1 as well to help indicate churn
for (i in 1:length(activeLogins2$firstlogindate)){
  myrows <- activeLogins2$startcol[i]:activeLogins2$endcol[i]
  postwk1rows <- (activeLogins2$endcol[i]+1):(activeLogins2$lastdatecol[i])
  activeLogins2$NumfirstWeekAL[i] <- sum(activeLogins2[i, myrows])
  activeLogins2$NumPostwk1AL[i] = sum(activeLogins2[i, postwk1rows])
} 


##====================================================================================
#  DEFINE CHURN AND ACTIVE CHURN
##====================================================================================
# Define churners as those that had ZERO active login days post wk 1 (churners=1, those with >1 AL = 0)
churnAL <- which(activeLogins2$NumPostwk1AL == 0) # index for churners
activeAL <- which(activeLogins2$NumPostwk1AL != 0) # index for churners
  
activeLogins2$churnAL <- 0
activeLogins2$churnAL[churnAL] <- 1  #index to which had activity

### ADD NumfirstWeekSM and Numpostwk1SM to loginInfo7
tempdf <- data.frame("userID"=profTimelineitem3$userID, "NumfirstWeekAL"=activeLogins2$NumfirstWeekAL, "NumfPostwk1AL" = activeLogins2$NumPostwk1AL, "churnAL" = activeLogins2$churnAL)
loginInfo7 <-  left_join(loginInfo7, tempdf, by = "userID")

## Add together total number of active sessions
loginInfo7$totalActiveLogins <- loginInfo7$NumfirstWeekAL + loginInfo7$NumfPostwk1AL
```


FEATURE #7: Subscription plan
Although at date of writing, this is unlikely to be an important feature in the model because there are low number of subscribers, BUT this is likely an important feature as more users become subscribers.
Additionally, the company can also use this as an outcome in the future to predict and define churn at this higher level. 
```{r}
## Introduce subscription plan information
subscriptionPlan2 <- data.frame("planNum" = subscriptionPlan$X_id, "plan" = subscriptionPlan$name, "planMonths" = subscriptionPlan$durationInMonths)

# can also include purchase date if interested
subscription2 <- data.frame("userID" = subscription$user, "planNum" = subscription$plan, "userType" = as.character("premium"))

length(which(subscription2$planNum==subscriptionPlan2$planNum[1])) # number of monthly plans
length(which(subscription2$planNum==subscriptionPlan2$planNum[2])) # number of annual plans

# add in which plan type 
monthlyPlans<- (which(subscription2$planNum==subscriptionPlan2$planNum[1])) # monthly plans
annualPlans <- (which(subscription2$planNum==subscriptionPlan2$planNum[2])) # annual plans
subscription2$planType <- ('')
subscription2$planType[monthlyPlans] <- as.character("monthly")
subscription2$planType[annualPlans] <- as.character("annual")
subscription2$planType <- as.factor(subscription2$planType)
subscription2$userType <- as.factor(subscription2$userType)

loginInfo7 <-  left_join(loginInfo7, subscription2, by = "userID")

#check levels
levels(loginInfo7$userType)
levels(loginInfo7$planType)

#add new factor level. i.e "free" or "none"
loginInfo7$userType <- factor(loginInfo7$userType, levels=c(levels(loginInfo7$userType), "free"))
loginInfo7$planType <- factor(loginInfo7$planType, levels=c(levels(loginInfo7$planType), "none"))

#convert all NA's to free or none
loginInfo7$userType[is.na(loginInfo7$userType)] <- "free"
loginInfo7$planType[is.na(loginInfo7$planType)] <- "none"

```

Final engineering of information into dataset to be written to .csv and imported into python for modeling
```{r}
### Binarize gender (0 = female, 1 = Male, else NA)
loginInfo7$sex <- NA
males <- which(loginInfo7$gender=="male")
females <- which(loginInfo7$gender=="female")
loginInfo7$sex[males] <- 1
loginInfo7$sex[females] <- 0

# ensure that churn is correctly measured
# Define churners as those that had more than 2 logins post week 1
churn <- which(loginInfo7$NumPostwk1Logins <= 1) # index for churners
loginInfo7$churn <- 0
loginInfo7$churn[churn] <- 1  #index to which had activity



# create data frame to export
modeldf <- data.frame("userID" = loginInfo7$userID, "dayswithApp" = loginInfo7$dayswithAppNumeric, "Wk1ActiveDays" = loginInfo7$Wk1ActiveDays, "NumfirstWeeklogins" = loginInfo7$NumfirstWeeklogins, "NumfirstWeekRuns" = loginInfo7$NumfirstWeekRuns, "NumfirstWeekSess" = loginInfo7$NumfirstWeekSess, "NumfirstWeekEnroll" = loginInfo7$NumfirstWeekEnroll, "NumfirstWeekSM" = loginInfo7$NumfirstWeekSM, "NumfirstWeekAL" = loginInfo7$NumfirstWeekAL, "planType" = loginInfo7$planType, "userType" = loginInfo7$userType, "churn" = loginInfo7$churn, "churnAL" = loginInfo7$churnAL, stringsAsFactors = TRUE)

# Replace any NA values with 0
modeldf$NumfirstWeekEnroll <- modeldf$NumfirstWeekEnroll%>% replace_na(0)
modeldf$NumfirstWeekRuns <- modeldf$NumfirstWeekRuns%>% replace_na(0)
modeldf$NumfirstWeekSess <- modeldf$NumfirstWeekSess%>% replace_na(0)
modeldf$NumfirstWeekSM <- modeldf$NumfirstWeekSM%>% replace_na(0)
modeldf$NumfirstWeekAL <- modeldf$NumfirstWeekAL%>% replace_na(0)
# churn measure, if they had more than 1 login in post week 1
# second churn measure, if they did not have an active login, replace na with 1 (they churned)
modeldf$churnAL <- modeldf$churnAL%>% replace_na(1)

modeldf$userTypeNumeric <- 0
modeldf$userTypeNumeric[which(modeldf$userType =="premium")] <- 1

 # Remove individuals that logged in >99 times **These are the coaches
 modeldf2 <- modeldf[which(modeldf$NumfirstWeeklogins < 100),]

 ### Creating a new churn for y
 ## ACTIVE, engaged users = 0, passive logins = 1, churners = 2
 modeldf2$churnlevels <- as.factor(modeldf2$churn + modeldf2$churnAL)

 # get the count for each level of churn
  modeldf2 %>% group_by(churnAL) %>% tally()
 modeldf2 %>% group_by(churn) %>% tally()
 modeldf2 %>% group_by(churnlevels) %>% tally()

 # the following csv file is what will be read into the python script for random forest modeling
#write.csv(modeldf2, "modeldf_updatedmultiChurn.csv")

 # Individuals that had SOME engagement with the app based on 'non churners'
#wk1active <- modeldf2[which(modeldf2$NumfirstWeekAL>=1),]
#write.csv(wk1active, "wk1activeONLY.csv")
```
